Racial and Ethnic Differences in ETI: Outcomes

### 2024-05-10 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti_outcomes.qmd
plots only

---
title: "Racial and Ethnic Differences in ETI: Outcomes"
author: "Tim Vigers"
date: "today"
date-format: long
format:
  html:
    toc: true
    toc-depth: 5
    toc-float: true
    number-sections: true
    code-fold: true
    self-contained: true
    fig-cap-location: top
    page-layout: full
    theme:
      light: flatly
      dark: darkly
editor: source
---

@@

```{r data setup}
#| include: false
# Import
encounter <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_encountersMerged_Del1.csv", na.strings = "")
annual <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_Annualized_Del1.csv", na.strings = "")
demo <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_DemogCFDiag_Del1.csv", na.strings = "")
# Race variable
races <- list(
  "Race1" = "White",
  "Race2" = "Black or African American",
  "Race3" = "American Indian or Alaska Native",
  "Race4" = "Asian",
  "Race5" = "Native Hawaiian or Other Pacific Islander",
  "Race6" = "Some other race"
)
demo$Race <- apply(demo[, paste0("Race", 1:6)], 1, function(r) {
  w <- which(r == 1)
  if (length(w) > 1 | 6 %in% w) {
    return("Mixed/Other/Unknown Race")
  } else {
    return(as.character(races[w]))
  }
})
demo$Race[demo$Hispanicrace == 1] <- "Hispanic or Latino"
demo$Race[is.na(demo$Hispanicrace)] <- "Mixed/Other/Unknown Race"
demo$Race <- factor(demo$Race,
  levels = c(
    "White", "Black or African American", "Hispanic or Latino", "Asian",
    "American Indian or Alaska Native",
    "Native Hawaiian or Other Pacific Islander", "Mixed/Other/Unknown Race"
  ),
  labels = c(
    "Non-Hispanic White", "Black or African American", "Hispanic or Latino", "Asian",
    "Other", "Other", "Other"
  )
)
# Insurance
ins_vars <- c(paste0("insurance", 1:7), "noinsurance")
ins <- c(
  "Health Insurance Policy (e.g. Private Insurance)", "Medicare", "Medicaid",
  "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
  "TriCare or other military health plan",
  "Indian Health Service", "Other", "No Insurance"
)
annual$Insurance <- apply(annual[, ins_vars], 1, function(r) {
  w <- which(r == 1)
  if (length(w) == 0) {
    return(NA)
  } else if (length(w) > 1) {
    return("More than one")
  } else {
    return(ins[w])
  }
})
annual$Insurance <- factor(annual$Insurance,
  levels = c(
    "Health Insurance Policy (e.g. Private Insurance)",
    "Indian Health Service",
    "Medicaid",
    "Medicare",
    "More than one",
    "No Insurance",
    "Other",
    "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
    "TriCare or other military health plan"
  ),
  labels = c(
    "Private", "Public", "Public", "Public", "More Than One", "No Insurance",
    "No Insurance", "Public", "Tricare"
  )
)
# Merge
encounter <- left_join(
  encounter,
  demo %>% select(eDWID, Race, Modulator_trikafta_first_date)
)
encounter <- left_join(encounter,
  annual %>% select(eDWID, ReviewYear, Insurance),
  by = join_by(eDWID, reviewyear == ReviewYear)
)
# Age group
encounter$age_group <- cut(encounter$encounterage, breaks = c(6, 12, 18, 30, Inf), right = F)
# Days between encounter and first ETI
encounter$Modulator_trikafta_first_date <-
  mdy(encounter$Modulator_trikafta_first_date)
encounter$encounterdate <- mdy(encounter$encounterdate)
encounter$Days <- as.numeric(encounter$encounterdate - encounter$Modulator_trikafta_first_date)
# Separate cleaned set
cleaned <- encounter %>% filter(gli_fev1_ppred_rn <= 200, !is.na(Days))
# List of outcomes
continuous_outcomes <- c("gli_fev1_ppred_rn", "gli_fvc_ppred_rn", "gli_fev1fvc_ppred_rn")
```

```{r functions}
# Plots
loess_plot <- function(df = cleaned, outcome, facet = F) {
  p <- ggplot(df, aes(x = Days, y = !!sym(outcome), color = Race)) +
    geom_point(shape = ".", aes(x = Days, y = !!sym(outcome)), inherit.aes = F) +
    geom_smooth() +
    theme_classic()
  if (facet) {
    p <- p + facet_wrap(~Race)
  }
  return(p)
}
```

@@

```{r results='asis'}
#| warning: false
invisible(lapply(continuous_outcomes, function(o) {
  cat("\n")
  cat("##", o)
  cat("\n")
  print(loess_plot(outcome = o))
  print(loess_plot(outcome = o, facet = T))
}))
```

@@

@@

```{r}
#| cache: true
set.seed(1017)
model <- list(
  gli_fev1_ppred_rn ~ 1, # intercept
  gli_fev1_ppred_rn ~ 1 ~ 0 + Days # joined slope
)
fit <- mcp(model, data = cleaned, chains = 4, cores = 4, iter = 1000, adapt = 0) # Increase iter and adapt later, because it will take forever to run
```

@@

## Frequentist models

First, a Gaussian mixed model is fit with random intercept for patient ID. Then, the `segmented.lme` function is applied to this mixed model, with 0 as the starting value for finding the optimal breakpoint.

```{r}
#| cache: true
# Fit lme
mod1 <- lme(gli_fev1_ppred_rn ~ Days * Race + Insurance,
  random = ~ 1 | eDWID,
  data = cleaned, na.action = na.omit
)
# mod1_seg <- segmented.lme(mod1, ~Days,
#   psi = 0,
#   random = list(eDWID = pdDiag(~ 1 + Days + U)), # ind. random eff. (changepoint fixed)
#   control = seg.control(n.boot = 0, display = TRUE)
# )
mod1_seg <- segmented.lme(mod1, ~Days,
  psi = 0,
  random = list(eDWID = pdDiag(~ 1 + Days + U + G0)), # ind. random eff. (in the changepoint too)
  control = seg.control(n.boot = 0, display = TRUE)
)
```

@@

@@

>>>>

---
title: "Racial and Ethnic Differences in ETI: Outcomes"
author: "Tim Vigers"
date: "today"
date-format: long
format:
  html:
    toc: true
    toc-depth: 5
    toc-float: true
    number-sections: true
    code-fold: true
    self-contained: true
    fig-cap-location: top
    fig-height: 10
    fig-width: 10
    page-layout: full
    theme:
      light: flatly
      dark: darkly
editor: source
---

@@

```{r data setup}
#| include: false
# Import
encounter <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_encountersMerged_Del1.csv", na.strings = "")
annual <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_Annualized_Del1.csv", na.strings = "")
demo <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_DemogCFDiag_Del1.csv", na.strings = "")
# Race variable
races <- list(
  "Race1" = "White",
  "Race2" = "Black or African American",
  "Race3" = "American Indian or Alaska Native",
  "Race4" = "Asian",
  "Race5" = "Native Hawaiian or Other Pacific Islander",
  "Race6" = "Some other race"
)
demo$Race <- apply(demo[, paste0("Race", 1:6)], 1, function(r) {
  w <- which(r == 1)
  if (length(w) > 1 | 6 %in% w) {
    return("Mixed/Other/Unknown Race")
  } else {
    return(as.character(races[w]))
  }
})
demo$Race[demo$Hispanicrace == 1] <- "Hispanic or Latino"
demo$Race[is.na(demo$Hispanicrace)] <- "Mixed/Other/Unknown Race"
demo$Race <- factor(demo$Race,
  levels = c(
    "White", "Black or African American", "Hispanic or Latino", "Asian",
    "American Indian or Alaska Native",
    "Native Hawaiian or Other Pacific Islander", "Mixed/Other/Unknown Race"
  ),
  labels = c(
    "Non-Hispanic White", "Black or African American", "Hispanic or Latino", "Asian",
    "Other", "Other", "Other"
  )
)
# Insurance
ins_vars <- c(paste0("insurance", 1:7), "noinsurance")
ins <- c(
  "Health Insurance Policy (e.g. Private Insurance)", "Medicare", "Medicaid",
  "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
  "TriCare or other military health plan",
  "Indian Health Service", "Other", "No Insurance"
)
annual$Insurance <- apply(annual[, ins_vars], 1, function(r) {
  w <- which(r == 1)
  if (length(w) == 0) {
    return(NA)
  } else if (length(w) > 1) {
    return("More than one")
  } else {
    return(ins[w])
  }
})
annual$Insurance <- factor(annual$Insurance,
  levels = c(
    "Health Insurance Policy (e.g. Private Insurance)",
    "Indian Health Service",
    "Medicaid",
    "Medicare",
    "More than one",
    "No Insurance",
    "Other",
    "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
    "TriCare or other military health plan"
  ),
  labels = c(
    "Private", "Public", "Public", "Public", "More Than One", "No Insurance",
    "No Insurance", "Public", "Tricare"
  )
)
# Merge
encounter <- left_join(
  encounter,
  demo %>% select(eDWID, Race, Modulator_trikafta_first_date)
)
encounter <- left_join(encounter,
  annual %>% select(eDWID, ReviewYear, Insurance),
  by = join_by(eDWID, reviewyear == ReviewYear)
)
# Age group
encounter$age_group <- cut(encounter$encounterage, breaks = c(6, 12, 18, 30, Inf), right = F)
# Days between encounter and first ETI
encounter$Modulator_trikafta_first_date <-
  mdy(encounter$Modulator_trikafta_first_date)
encounter$encounterdate <- mdy(encounter$encounterdate)
encounter$Days <- as.numeric(encounter$encounterdate - encounter$Modulator_trikafta_first_date)
# Separate cleaned set
cleaned <- encounter %>% filter(gli_fev1_ppred_rn <= 200, !is.na(Days))
# List of outcomes
continuous_outcomes <- c(
  "gli_fev1_ppred_rn", "gli_fvc_ppred_rn", "gli_fev1fvc_ppred_rn",
  "bmipercentile", "heightpercentile", "weightpercentile"
)
binary_outcomes <- c(
  "staphylococcus_aureus", "haemophilus_influenzae", "pseudomonasaeruginosa",
  "burkho_complex"
)
```

```{r functions}
# Plots
loess_plot <- function(df = cleaned, outcome, facet = F) {
  p <- ggplot(df, aes(x = Days, y = !!sym(outcome), color = Race)) +
    geom_point(
      shape = ".", aes(x = Days, y = !!sym(outcome)), inherit.aes = F,
      alpha = 0.1
    ) +
    geom_smooth() +
    theme_classic()
  if (facet) {
    p <- p + facet_wrap(~Race)
  }
  return(p)
}
# Basic segmented model
basic_segmented <- function(df = cleaned, outcome, n_change_points = NULL,
                            change_points = NULL) {
  # Formula
  f <- as.formula(paste0(outcome, "~ Days+Race + age_group + Insurance"))
  # Basic linear model
  lmod <- lm(f, data = df)
  # Segmented model
  lmod_seg <- segmented(lmod, seg.Z = ~Days, psi = 0, control = seg.control(it.max = 0))
  # Plot
  d <- lmod$model
  d$pred <- fitted(lmod_seg)
}
```

@@

```{r results='asis'}
#| warning: false
invisible(lapply(continuous_outcomes, function(o) {
  cat("\n")
  cat("\n")
  cat("##", o)
  cat("\n")
  cat("\n")
  print(loess_plot(outcome = o))
  print(loess_plot(outcome = o, facet = T))
}))
```

@@

## FEV1

Based on visual inspection of the LOESS plots, I expect there to be a change in slope at around 0 and maybe a second one at around 1 year.

@@

```{r}
#| cache: true
set.seed(1017)
model <- list(
  gli_fev1_ppred_rn ~ 1, # intercept
  gli_fev1_ppred_rn ~ 1 ~ 0 + Days # joined slope
)
fit <- mcp(model, data = cleaned, chains = 4, cores = 4)
```

@@

<!-- ## Frequentist models -->

<!-- First, a Gaussian mixed model is fit with random intercept for patient ID and random slope for each patient. Then, the `segmented.lme` function is applied to this mixed model, with 0 as the starting value for finding the optimal breakpoint. The changepoint is also treated as a random effect and allowed to vary within a person. -->

<!-- ### FEV1 -->

<!-- ```{r} -->
<!-- #| cache: true -->
<!-- # Fit initial model -->
<!-- ri_rs_mod_fev1 <- lme(gli_fev1_ppred_rn ~ Race * Days + age_group + Insurance, -->
<!--   random = ~ 1 + Days | eDWID, data = cleaned, na.action = na.omit -->
<!-- ) -->
<!-- # Find change point -->
<!-- ri_rs_mod_seg_fev <- segmented.lme(ri_rs_mod_fev1, ~Days, psi = 0, -->
<!--   random = list(eDWID = pdDiag(~ 1 + Days + U + G0)), # ind. random eff. (in the changepoint too) -->
<!--   control = seg.control(n.boot = 0, display = F) -->
<!-- ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- plot(ri_rs_mod_seg_fev1) -->
<!-- ``` -->

@@

5. Do the linearity assumptions seem accurate to you?

6. Why do you think the BMI, height, etc. percentiles look so strange? Do they normally look like this?

@@

- Is the second change point due to who is being collected after one year out?
  - 
- Kam will

< 20% predicted and >=150 (Kam)

Look at 3 years before and two after

2 years for 12 and up
1.5 years for 6
